### 2025-02-24 paper/sections/other/training.qmd

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Explanations, Explainable
AI, Representation Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed plausible and
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning plausible explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and plausible explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver plausible
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

\section{Introduction}\label{introduction}

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning these
representations from scratch, guided by narrow objectives such as
predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
Bengio, and Courville 2016}). Modern advances in computing have made it
possible to provide such models with ever greater degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so they also learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
showed that artificial neural networks (ANN) are sensitive to
adversarial examples: counterfactuals of model inputs that yield vastly
different model predictions despite being ``imperceptible'' in that they
are semantically indifferent from their factual counterparts. Despite
partially effective mitigation strategies such as \textbf{adversarial
training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
  clarity: we follow standard ML convention in using ``degrees of
  freedom'' to refer to the number of parameters estimated from data.}.
Based purely on predictive performance, these solutions may seem to
provide compelling explanations for the data, when in fact they are
based on purely associative, semantically meaningless patterns. This
poses two related challenges: firstly, it makes these models inherently
opaque, since humans cannot simply interpret what type of explanation
the complex learned representations correspond to; secondly, even if we
could resolve the first challenge, it is not obvious how to mitigate
models from learning representations that correspond to meaningless and
implausible explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems (ADMS) involving
opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
XAI tools can aide us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g.~``loan application
rejected'') into positive ones (``application accepted''). Wachter,
Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) propose
\textbf{counterfactual explanations} as an effective approach to achieve
this: they explain how factual inputs need to change in order for some
fitted model to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed \textbf{plausible} by
humans (see Definition~\ref{def-explainability}). Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both plausible and valid for any given
model, independent of whether the learned representations are also
compatible with implausible explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with plausible explanations. Our
contributions are as follows:

\begin{itemize}
\tightlist
\item
  We discuss existing related work on improving models and consolidate
  it through the lens of counterfactual explanations
  (Section~\ref{sec-lit}).
\item
  We present our proposed methodological framework that leverages
  faithful counterfactual explanations during the training phase of
  models to achieve the explainability objective
  (Section~\ref{sec-method}).
\item
  Through extensive experiments we demonstrate the counterfactual
  training improve model explainability while maintaining high
  predictive performance. We run ablation studies and grid searches to
  understand how the underlying model components and hyperparameters
  affect outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite limitations of our approach discussed in
Section~\ref{sec-discussion}, we conclude that counterfactual training
provides a practical framework for researchers and practitioners
interested in making opaque models more trustworthy
Section~\ref{sec-conclusion}. We also believe that this work serves as
an opportunity for XAI researchers to reevaluate the premise of
improving XAI tools without improving models.

\section{Related Literature}\label{sec-lit}

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

\subsection{Adversarial Examples are Counterfactual
Explanations}\label{adversarial-examples-are-counterfactual-explanations}

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization. Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function.

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation: Wu et
al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{Polyjuice}, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through
\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
tasks. Balashankar et al.
(\citeproc{ref-balashankar2023improving}{2023}) also use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves up to 20\%. Finally, Luu
and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
Counterfactual Adversarial Training (CAT), which also aims at improving
generalization and robustness of language models. Specifically, they
propose to proceed as follows: firstly, they identify training samples
that are subject to high predictive uncertainty; secondly, they generate
counterfactual explanations for those samples; and, finally, they
fine-tune the given language model on the augmented dataset that
includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations (CE) and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
case for jointly studying the opaqueness and robustness problem in
representation learning. Formally, AE can be seen as the subset of CE,
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
CE and AE are equivalent under certain conditions and derive theoretical
upper bounds on the distances between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
propose a way to train models that are guaranteed to provide recourse
for individuals to move from an adverse outcome to some positive target
class with high probability. The approach proposed by Ross, Lakkaraju,
and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
adversarial training, where in this context susceptibility to targeted
adversarial examples for the positive class is explicitly induced. The
proposed method allows for imposing a set of actionability constraints
ex-ante: for example, users can specify that certain features
(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
to propose an end-to-end training pipeline that includes counterfactual
explanations as part of the training procedure. In particular, they
propose a specific network architecture that includes a predictor and CE
generator network, where the parameters of the CE generator network are
learnable. Counterfactuals are generated during each training iteration
and fed back to the predictor network. In contrast to Guo, Nguyen, and
Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
restrictions on the neural network architecture at all.

\subsection{Beyond Robustness}\label{beyond-robustness}

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
be improved through model averaging and refined model objectives. The
authors propose a way to generate counterfactuals that are maximally
\textbf{faithful} to the model in that they are consistent with what the
model has learned about the underlying data. Formally, they rely on
tools from energy-based modelling to minimize the divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by the model. Their proposed counterfactual explainer,
\emph{ECCCo}, yields plausible explanations if and only if the
underlying model has learned representations that align with them. They
find that both deep ensembles
(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
and Blundell 2017}) and joint energy-based models (JEMs)
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work by Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating plausible (``interpretable'') counterfactual explanations is
almost trivial for deep ensembles that have also undergone adversarial
training. The case for JEMs is even clearer: they involve a hybrid
objective that induces both high predictive performance and generative
capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
is closely related to the idea of aligning models with plausible
explanations and has inspired our proposed counterfactual training
objective, as we explain in Section~\ref{sec-method}.

\section{Counterfactual Training}\label{sec-method}

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

\begin{definition}[Model
Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}

Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
supervised classification model that maps from the \(D\)-dimensional
input space \(\mathcal{X}\) to representations
\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
space \(\mathcal{Y}\). Assume that for any given input-output pair
\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
where \(\arg\max_y{\mathbf{y}^{+}}=y^+\) and \(y^+\) denotes the index
of the target class.

We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
that faithfully generated counterfactuals are plausible (i.e.~consistent
with the data) and actionable. Formally, we define these properties as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  (Plausibility)
  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
\item
  (Actionability) Permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

We consider counterfactuals as faithful to the extent that they are
consistent with what the model has learned about the input data. Let
\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
over inputs, then formally:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  (Faithfulness)
  \(\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is defined as above.
\end{enumerate}

\end{definition}

The definitions of faithfulness and plausibility in
Definition~\ref{def-explainability} are the same as in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
Actionability constraints in Definition~\ref{def-explainability} vary
and depend on the context in which \(\mathbf{M}_\theta\) is deployed. In
this work, we focus on domain and mutability constraints for individual
features \(x_d\) for \(d=1,...,D\). We limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

\subsection{Our Proposed Objective}\label{our-proposed-objective}

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
let \(T\) vary and consider the counterfactual search as converged as
soon as the predicted probability for the target class has reached a
pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
  detailed background information on gradient-based counterfactual
  search and convergence see Section~\ref{sec-app-ce}.}.

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose to leverage
counterfactuals in the following objective,

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g.~cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: 1) the contrastive
divergence, \(\text{div}(\cdot)\), between counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

\subsubsection{Directly Inducing Explainability through Contrastive
Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}

Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
any classifier can be re-interpreted as a joint energy-based model (JEM)
that learns to discriminate output classes conditional on inputs and
generate inputs. They show that JEMs can be trained to perform well at
both tasks by directly maximizing the joint log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
contrastive divergence between samples drawn from
\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
\(p(\mathbf{x})\).

A key empirical finding in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.
Specifically, we have

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class. We generate samples
\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
that target over \(T\) iterations using a gradient-based counterfactual
generator. This is similar to how conditional sampling is used to draw
from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}).

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
counterfactuals get more plausible (Definition~\ref{def-explainability})
over the cause of training, these two opposing effects gradually balance
each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. Typically, these methods facilitate the imposition of domain
and mutability constraints, for example. In principle, any existing
approach for generating counterfactual explanations is viable, so long
as it does not violate the faithfulness condition. Like JEMs
(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
training can be considered as a form of contrastive representation
learning.

\subsubsection{Indirectly Inducing Explainability through Adversarial
Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or
adversarial examples
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility with respect to the exact choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form to use, though likely not the only reasonable choice, is
inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation}

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
magnitude of the perturbation to any individual feature is at most
\(\varepsilon\). This is closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage counterfactual
explanations during training and get adversarial examples, essentially
for free.

\subsection{Encoding Actionability Constraints}\label{sec-constraints}

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
In this context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step resulted in updated feature values that were out-of-domain.
Mutability constraints can similarly be enforced by setting partial
derivatives to zero to ensure that features are only mutated in the
allowed direction, if at all.

>>>>

## Training Details {#sec-app-training}
